Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-12-29 docs/ms_sageAbundance.Rmd
add PBA edits

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
We outlined a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which inspires confidence in our forecasts.
More generally, the approach we described is applicable to any spatially-explicit time series of plant population abundance and opens the door to forecasts of plant populations at the landscape-scale.

### Sagebrush response to climate and climate change
The climate effects estimated in our model, and detected at a spatial resolution of 30 meters, are consistent with individual-level responses of sagebrush to climate-related variables.
At the level of individual plants, research shows that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and warmer spring temperatures may enhance sagebrush growth in colder climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
Mirroring those individual-level responses, posterior means for all climate effects in our cover change model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig. 3).
The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, the mean estimates for the climate effects are relatively small on the standardized scale (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
But, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration.
More likely is that aggregated estimates of plant abundance, like percent cover, mask interannual variability at the level of the individual plant.
Such averaging inevitably reduces interannual variability, which also makes detecting drivers of internanual variability more difficult.
Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.

Under projected climate, we forecast small positive changes in sagebrush cover in this area for all RCP scenarios (Figs. 5,6).
Our forecasts are a direct result of the interaction between the estimated effect size for each cliamte covariate and the amount of change in those covariates projected in the future.
First, precipitation two years before a given year-to-year transition was the most important effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2).
Second, mean spring temperature had a positive, but small, effect on sagebrush cover changes, while the projected change in temperature was large (Table 1, Fig. 2).
Lastly, the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the increase in spatial heterogeity we forecast: the distribution of sagebrush cover expands rather than just shifting (Fig. D1).
This effect is seen in the spatial forecasts where low cover pixels are constrained to low cover while high cover pixels increase (Fig. 5).
Had we not explicitly accounted for spatial-dependence in our model, this result would be missed.

While we forecast an increase in sagebrush cover at our study area, SDM studies tend to project declines in sagebrush abundance in the future due to temperature increases [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
Why do our results differ?
There are many potential reasons, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in the cold range of the sagebrush distrubtion is the most likely explanation.
The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
The generally positive effect of precipitation we found is consistent with findings from across the sagebrush distribution [@Schlaepfer2012; @Schlaepfer2015].
However, the effect of temperature may vary depending on location.
In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
Since our study area lies at the cold end of the sagebrush distribution, the weak positive response to temperature detected by our statistical model (Fig. 3) and carried through to our forecasts (Figs. 5,6) likely represents the positive effect of earlier snowmelt.

Also in contrast to our results, a previous analysis of a different subset of the remote sensing data set we used made projections of sagebrush decline, on average, in the future [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours reflects a difference in the climate projections used for forecasting future changes rather than differences in our inference about how climate drives sagebrush cover change.
First, Homer et al. (2015) used downscaled IPCC projections that allowed for spatial variation in future climate, unlike our analysis here.
Therefore, Homer et al. (2015) did predict some pixel-level increases in sagebrush cover since each pixel was subject to its own regression between climate and cover.
Second, like our study, Homer et al. (2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the climate scenario they chose resulted in a mean decrease in precipitation.
Thus, the decline in precipitation led to a decline in sagebrush cover.
Our current work shows that when multiple climate projections are considered, on average sagebrush cover is forecast to increase, but decreases are not outside the realm of our forecasts when all sources of uncertainty are considered (Fig. 6).
Lastly, Homer et al.'s (2015) analysis relied on pixel-level regressions of cover through time against climate covariates.
Our current approach is fundamentally different in that we specifically model the impact of weather on year-to-year changes in sagebrush cover.
Thus, our model captures the essential interaction point between populations and the interannual variation in weather that drives population dynamics.

We presented projections for sagebrush cover at the end of this century, but an important feature of our approach is that it could also produce short-term forecasts for sagebrush cover.
For example, we could forecast the effects of a multi-year regional drought on this species [@Debinski2010].
Validating spatial population models against short-term predictions would give ecological forecasters a way to assess the long-term forecasts produced by the model.
It would also provide opportunities for refinement of the model itself.
This cycle of prediction, validation, and refinement is missing from most population-level forecasts of the effects of climate change currently produced.

@@

Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
We focused the native spatial resolution of the remote sensing data ($30\times30$ meters).
This constrained the extent that we could reasonably model because of the computational challenges in using a massive spatial random effect. 
Even with our dimension reduction technique modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
To overcome this, future investigators modeling a larger spatial extent may need to aggregate their original remote-sensing time series data to a coarse spatial resoluation. This would allow them to model a much greater spatial extent with a similar number of knots and a similar amount of computation time.
A coarser scale model would lose some fine-scale detail but could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation.
However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger and larger pixels.

Our spatial extent and resolution also impacted our incorporation of climate covariates.
We did not downscale Daymet data to match the spatial resolution of the sagebrush data.
Thus, in each year all pixels share the same climate covariates.
This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).
We also did not allow for the possibility that different parts of our study area would respond to climate in different ways.
This would require spatially-varying climate effects that substantially increase computational time.
However, in some case, for example over very large regions, allowing the effect of climate to vary over space will be important.
@Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
Future studies could examine lower spatial resolutions and larger spatial extents to facilitate the use of spatially-explicit and spatially-varying climate covariates that match the resolution of the percent cover data.
It is also possible to include relevant abiotic data layers as predictors when fitting models at larger spatial extents where variation in spatial covariates like soil type, slope, and aspect may play more important roles than we observed in our study area.

As mentioned above, one of the main differences between our current work and that of Homer et al. (2015) is our choice of climate projections.
Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013]. There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
Extensions of these ideas are the subject of ongoing research.

>>>>

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
Our approach allowed us to fit a dynamic population model and to make spatially-explicit forecasts of sagebrush response to climate change at the landscape scale.
The climate effects detected by our statistical model are in agreement with previous findings at the individual and landscape scale, which imspires some confidence in our forecasts.
More generally, the approach we described is applicable to any spatially-explicit time series of plant population abundance and opens the door to forecasts of plant populations at the landscape-scale.

### Sagebrush response to climate and climate change
The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation one year before a cover transition (*ppt1*, Fig. 3).
The amount of precipitation two years before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, the mean estimates for the climate effects are relatively weak (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration (e.g. REFERENCE????).
More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual 
variability. Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.

Under projected climate, we forecast modest increases in sagebrush cover for all RCP scenarios (Figs. 5,6).
Our forecasts reflect both the estimated effect size for each climate covariate and the amount of change in those covariates projected by the GCMs.
Precipitation two years before a given year-to-year transition was the strongest standardized effect (Fig. 3), but precipitation is projected to increase only moderately (Table 1, Fig. 2) and the negative effect of precipitation two years before a cover transition (*ppt1*) had an offsetting effect.
In contrast, mean spring temperature had a weak positive effect on sagebrush cover changes, but the projected temperature increase is large (Table 1, Fig. 2).
An interesting consequence of explicitly modeling the effect of space (through $\boldsymbol{\eta}$) is the forecast increase in spatial heterogeity (Fig. D1).
Our model projects little change in low cover pixels but substantial increases in the cover of high cover pixels (Fig. 5). Had we not explicitly accounted for spatial-dependence in our model, we would have missed this result.

While we forecast an increase in sagebrush cover at our study area, SDM studies typically project dramatic declines in climate suitability for sagebrush with warming [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
There are many potential explanations for this apparent contrast, ranging from the type of model used to the particular climate covariates considered, but the location of our study area in a cold portion of sagebrush's geographic distribution may be the best.
The response of plant species to weather varies along climatic gradients [e.g., @Clark2011; @Vanderwel2013b], and sagebrush are especially sensitive to the timing of snowmelt since their growth depends on recharge of deep soil water [@Schlaepfer2012; @Schlaepfer2014].
In warmer parts of the sagebrush range, earlier snowmelt is detrimental to growth and survival [@Pechanec1937; @Schlaepfer2011; @Germino2014].
In colder regions, earlier snowmelt due to temperature increases can lengthen the growing season and increase sagebrush occurence and cover [@Schlaepfer2012; @Schlaepfer2014].
Since our study area lies at the cold extreme of the sagebrush range, the weak positive response to temperature that we estimated (Fig. 3) and carried through to our forecasts (Figs. 5,6) likely represents the positive effect of earlier snowmelt.

A previous analysis of a different subset of the remote sensing data set we used also came to a different conclusion, projecting future sagebrush decline [@Homer2015].
The discrepancy between Homer et al.'s (2015) results and ours primarily reflects a difference in the climate projections used for projecting future changes rather than differences in our inference about responses to historical variation in weather.
Consistent with our study, Homer et al. (2015) found a generally positive relationship between pixel-level sagebrush cover and precipitation, but the future climate scenario they chose resulted in a mean decrease in precipitation, causing a predicted decline in sagebrush cover.
A second difference is that Homer et al.'s (2015) analysis relied on regressions of decadal trends in sagebrush cover against decadal trends in climate at the level of individual pixels.
Our current approach is fundamentally different in that we specifically model the impact of interannual variation in weather on year-to-year changes in sagebrush cover.
Thus, our model takes advantage of the additional information contained within short-term responses to climate fluctuations.

We projected sagebrush cover to the end of this century, but an important feature of our approach is that it could also produce short-term forecasts.
For example, we could forecast the effects of a multi-year regional drought on sagebrush cover [@Debinski2010]. Validating spatial population models against short-term predictions would give ecological forecasters a way to assess and improve the performance of their models, which would greatly increase our confidence in long-term forecasts.
This cycle of prediction, validation, and refinement is missing from most currently available population-level forecasts of the effects of climate change .

@@

Several *a priori* modeling decisions determined the spatial extent and resolution of our results. 
We retained the native spatial resolution of the remote sensing data ($30\times30$ meters).
This constrained the extent that we could reasonably model because of the computational challenges in estimating spatial random effects. 
Even with our dimension reduction technique, modeling a larger area at this resolution would require a greater number of spatial knots, and computation time would increase substantially [@Wikle2010].
To model a larger spatial extent, we could aggregate the original remote-sensing time series data to a coarser spatial resoluation. This would allow us to model a much greater spatial extent with a similar number of knots and a similar computation time.
While a coarser scale model would lose some fine-scale detail, it could be applied to a much larger area, potentially gaining some strength in estimating climate effects by spanning a greater range of climate variation. However, gains made by incorporating greater regional variability by modeling at a coarser resolution could be offset by the loss of information inherent when aggregating plant responses into larger and larger pixels.

Our spatial extent and resolution also affected our use of climate covariates.
We did not downscale Daymet data to match the spatial resolution of the sagebrush data, meaning that in each year all pixels share the same climate covariates.
This is a potential limitation of our study, and could explain the weak effect of climate covariates that we observed (Fig. 3).
We also did not allow different portions of our study area to respond to climate in different ways.
Doing so would require spatially-varying climate effects and a substantial increase in computational time.
However, in future applications it will be important to allow climate effects to vary over space.
@Conn2015 provide examples of how such spatiotemporal interactions can be included in abundance models.
We might expect climate effects to interact with spatial covariates such as soil type, slope, and aspect. In our relative small study area, we did not observe important effects of these factors, but it is possible to include such abiotic data layers as predictors when fitting models at larger spatial extents.

The contrast between our conclusions and those of Homer et al. (2015), mentioned above, emphasizes the irole of climate projections in ecological forecasting.
Weather projections from GCMs differ because each model incorporates processes and feedbacks in different ways [@Flato2013].
These structural differences among GMCs can lead to divergent weather forecasts for specific regions, even if the GCMs converge on projections at the global scale [@Flato2013]. There is no clear answer to what climate model to use when making ecological forecasts at the regional scale.
Much like weather forecasters [@Bauer2015], ecological forecasters may need to apply expert judgement to choose climate projections.
If a certain GCM tends to better predict weather in a specific study area [e.g., @Rupp2013], then that GCM may be given more weight than others.
Extensions of these ideas are the subject of ongoing research.
